---
title: "Pricing the Pandemic"
subtitle: "<span style='font-size: 1.2em; font-style: italic'>How COVID-19 Events Moved Markets</span>"
author: 
  - name: "Tuan Tran & Marsya Amanda  \nMentored by Dr. Beilei Bian"
format:
  revealjs:
    theme: serif
    title-slide-attributes:
      data-align: "left"
---

# Motivation & Background

## 

::: {.callout-important appearance="simple"}
### Motivation

-   Investigate market perception of pharmaceutical innovation and public health news.

-   Learn and apply event studies methods to study investor behaviour under uncertainty.
:::

::: {.callout-important appearance="simple"}
### Background

-   The widespread impact of COVID-19 events and vaccine innovation have been investigated in previous studies, but with less focus of the impact of these events on vaccine-producing companies themselves.

-   We focus on the three largest COVID-19 vaccine-producing companies by market capitalisation: Oxford-AstraZeneca, Janssen-Cilag International NV (sub-company of Johnson & Johnson), and Pfizer-BioNTech.

-   Pfizer-BioNTech was the COVID-19 vaccine to be approved by the United States Food and Drugs Association.
:::

# Methods

## Data Collection

-   **Tools:** `yfR` (Yahoo Finance) R package.

-   **Time period:** January 30, 2015 to November 8, 2025

-   **Stocks:** Pfizer-BioNTech `(PFE)`, Oxford-AstraZeneca `(AZN)`, Johnson & Johnson `(JNJ)`

-   **Foreign exchange rates:** GBP to USD exchange rate `(GBPUSD)`

-   **Market & risk-free rates:** S&P 500 `(GSPC)`, 10-year US treasury yields `(TNX)`

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false
library(dplyr)
library(ggplot2)
library(tidyr)
library(janitor)
library(patchwork)

# all results
company_results <- readRDS("../output/company-specific-results.rds")

# 3 largest in market cap
file_paths <- c("../data/JNJ_yfR_2015-01-30_2025-09-08.rds",
                "../data/rds/AZN_USD_2015-01-30_2025-09-08.rds", 
                "../data/PFE_yfR_2015-01-30_2025-09-08.rds")
azn <- readRDS(file_paths[2])
jnj <- readRDS(file_paths[1])
pfe <- readRDS(file_paths[3])

# Function
cumret <- function(file_path) {
  # Read data
  df <- readRDS(file_path)

  price_close <- df$price_close
  ret <- price_close / dplyr::lag(price_close) -1
  cumret <- cumprod(1 + coalesce(ret, 0)) - 1

  # Preprocess
  df <- df %>%
    group_by(ref_date) %>%
    select(contains("ticker"), ref_date, price_close, volume, price_adjusted) %>%
    mutate(
      price_close = as.numeric(price_close),
      price_adjusted = as.numeric(price_adjusted),
      volume = as.numeric(volume)
    ) %>%
    arrange(ref_date) %>%
    mutate(
      ret = !!ret,
      cumret = !!cumret)
  
  min_cumret <- min(df$cumret)
  ticker_col <- names(df)[grepl("ticker", names(df))]
  ticker_name <- df[[ticker_col[1]]][1]
  
  # Return useful objects
  return(list(
    data = df,
    ret = ret,
    cumret = cumret
  ))
}

# combine data sets for stacked line chart
jnj_mod <- as.data.frame(cumret(file_paths[1])[[1]])
azn_mod <- as.data.frame(cumret(file_paths[2])[[1]])
pfe_mod <- as.data.frame(cumret(file_paths[3])[[1]])

combined_cumret <- bind_rows(jnj_mod, azn_mod, pfe_mod)
```

# Exploratory Data Analysis (EDA)

## Data Structure

For each of `PFE`, `AZN`, `JNJ`, `GBPUSD` , and `^TNX`:

-   2666 observations of 7 variables: `ticker, ref_date, price_close, volume, price_adjusted, ret, cumret`

-   No missing values after filtering

-   `AZN` closing and adjusted price is converted to USD, with additional `ticker.x` and `ticker.y` columns.

## 

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: true

knitr::kable(head(azn), format = "markdown", caption = "Oxford-AstraZeneca (converted to USD)", digits = 2) %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = T,
                font_size = 13,
                position = "left")

knitr::kable(head(jnj), caption = "Johnson & Johnson", digits = 2)  %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = T,
                font_size = 13,
                position = "left")

knitr::kable(head(jnj), caption = "Pfizer-BioNTech", digits = 2)  %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = T,
                font_size = 13,
                position = "left")
```

## Market-wide events

::::: columns
::: {.column style="width "}
```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 7.5
#| paged-print: false

# plot stacked line chart - general events
gen_events <- c(as.Date("2020-01-30"),
                as.Date("2020-12-31"),
                as.Date("2023-05-05"))

zoomed_combined <- combined_cumret %>%
  filter(
    ref_date >= min(gen_events) - 300, 
    ref_date <= max(gen_events) + 300
  )

minc <- min(combined_cumret$cumret)

p_ge <- ggplot(zoomed_combined, aes(x = ref_date, y = cumret, group = ticker, color = ticker)) +
  geom_line() +
  geom_vline(
    xintercept = gen_events,
    colour = "gray30",
    linetype = "dashed"
    ) +
  annotate(
    "label",
    x = gen_events[1],
    y = 0.55,
    label = "PHEIC\nAnnounced",
    angle = 0,
    vjust = -0.5,
    size = 3,
    colour = "gray30"
  ) +
  annotate(
    "label",
    x = gen_events[2],
    y = 0.7,
    label = "First FDA\nApproval",
    angle = 0,
    vjust = -0.5,
    size = 3,
    colour = "gray30"
  ) +
  annotate(
    "label",
    x = gen_events[3],
    y = 0.8,
    label = "PHEIC\nRemoved",
    angle = 0,
    vjust = -0.5,
    size = 3,
    colour = "gray30"
  ) +
  labs(
    title = "Cumulative Returns of JNJ, AZN, and PFE",
    colour = "Ticker",
    x = "Date",
    y = "Cumulative returns since 2015-01-30"
  ) +
  theme_linedraw()
p_ge

```
:::

::: {.column style="width "}
-   obs1

-   obs2
:::
:::::

## Company-wide events

::::: columns
::: column
```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 6
#| paged-print: false
combined_data <- bind_rows(
  jnj_mod |> mutate(company = "JNJ"),
  azn_mod |> mutate(company = "AZN"),
  pfe_mod |> mutate(company = "PFE")
)

event_data <- tribble(
  ~company, ~date, ~label,
  "JNJ", as.Date("2020-07-10"), "Phase 1/2\nTrial Start",
  "JNJ", as.Date("2021-02-27"), "FDA EUA",
  "JNJ", as.Date("2021-04-13"), "Pause/Related Event",
  "AZN", as.Date("2020-04-23"), "World Trials\nCommence",
  "AZN", as.Date("2020-09-06"), "US Trials Halt",
  "AZN", as.Date("2020-10-23"), "US Trials Resume",
  "AZN", as.Date("2021-03-07"), "First Clots",
  "AZN", as.Date("2021-03-16"), "More Clots",
  "PFE", as.Date("2020-11-01"), "UK EUA",
  "PFE", as.Date("2021-08-23"), "FDA Approval",
  "PFE", as.Date("2021-10-29"), "FDA EUA\nAges 5–11"
)

min_event <- min(event_data$date)
max_event <- max(event_data$date)

# zoom window: 20 days before earliest event to 60 days after latest (you can adjust)
zoomed_data <- combined_data %>%
  filter(
    ref_date >= min_event - 20,
    ref_date <= max_event + 60
  )

# --- Faceted plot ---
ggplot(zoomed_data, aes(x = ref_date, y = cumret)) +
  geom_line() +
  geom_vline(
    data = event_data,
    aes(xintercept = date),
    colour = "gray30",
    linetype = "dashed"
  ) +
  ggrepel::geom_label_repel(
    data = event_data,
    aes(x = date, y = Inf, label = label),
    colour = "gray30",
    fill = "white",
    label.size = 0.2,
    direction = "y",
    size = 2.5,
    force = 2
  ) +
  facet_wrap(~ company, scales = "free_y", ncol = 1) +
  labs(
    title = "Cumulative Returns Around Major COVID-19 Vaccine Events",
    x = "Date",
    y = "Cumulative Returns (since 2015-01-30)"
  ) +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "gray90", color = NA),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.background = element_rect(fill = NA)
  )

```
:::

::: column
-   obs1

-   obs2

-   
:::
:::::

# Conducting Event Studies

## 

::: {.callout-note appearance="minimal"}
## Workflow

1.  **Define Pre-Event Window, Event Window, and Post-Event Window**

2.  **Compute Normal vs. Abnormal Returns**

3.  **Statistical Models for Expected Return**

    -   Market Model: $R_{i,t} = \alpha_i + \beta_i \cdot R_{m,t} + \epsilon_{i,t}$

    -   Market Adjusted Model: $R_{i,t} = R_{m,t} + \epsilon_{i,t}$

    -   Comparison Period Mean Adjusted Model: $R_{i,t} = \mu_i + \epsilon_{i,t}$

    -   Fama-French 3 Factor Model: $R_i - R_f = \alpha + \beta_1 \cdot (R_m - R_f) + \beta_2 \cdot \text{SMB} +\beta_3 \cdot \text{HML} + \epsilon$

4.  **Compute Cumulative Abnormal Returns over Event Window**

5.  **Conduct t-Test for Significance**

    $H_0: \text{Expected abnormal returns during the event window} = 0$
:::

## Which Model to Use?

# Results

## Market-Wide Events

-   result1

## Company-Wide Events

-   The **start of the global trials of the Oxford-AstraZeneca vaccine** resulted in a significant abnormal reaction in the AZN stock price (*p* = 0.0006) under the market model.

-   The **first** **emergency use authorisation (EUA) of the Johnson & Johnson vaccine** by the US FDA was a significant event causing abnormal stock prices (*p* = 0.0142) relative to the mean JNJ stock price in the last 250 days.

-   **Pfizer-BioNTech becoming the first vaccine to be approved by the US FDA** was significant (*p* = 0.0002) under the market model.

## Data

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

significant_results <- company_results %>%
  select(ends_with("_p")) %>%
  mutate(across(everything(), ~ ifelse(. < 0.05, ., NA_real_))) %>%
  filter(if_any(everything(), ~ !is.na(.))) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), "", .)))

p_results <- company_results %>%
  select(ends_with("_p"))

knitr::kable(p_results, digits = 4) %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "bordered"),
    full_width = T,
    font_size = 24
  ) %>%
  kableExtra::row_spec(
    row = 1,
    background = "lightblue"
  ) %>%
  kableExtra::column_spec(
    column = 2,
    bold = ifelse(p_results$MAM_p < 0.05, TRUE, FALSE)
  ) %>%
  kableExtra::column_spec(
    column = 3,
    bold = ifelse(p_results$CMA_p < 0.05, TRUE, FALSE)
  ) %>%
  kableExtra::column_spec(
    column = 4,
    bold = ifelse(p_results$MM_p < 0.05, TRUE, FALSE)
  ) %>%
  kableExtra::row_spec(
    row = 7,
    background = "lightblue"
  ) %>%
  kableExtra::row_spec(
    row = 10,
    background = "lightblue"
  )
```

## Conclusions

### Investor Behaviour

### Methodology

## References
