---
title: "Pricing the Pandemic"
subtitle: "<span style='font-size: 1.2em; font-style: italic'>How COVID-19 Events Moved Markets</span>"
author: 
  - name: "Tuan Tran & Marsya Amanda  \nMentored by Dr. Beilei Bian"
format:
  revealjs:
    theme: serif
    title-slide-attributes:
      data-align: "left"
  html:
    embed-resources: true
---

# Motivation & Background

## 

::: {.callout-important appearance="simple"}
### Motivation

-   Investigate market perception of pharmaceutical innovation and public health news.

-   Learn and apply event studies methods to study investor behaviour under uncertainty.
:::

::: {.callout-important appearance="simple"}
### Background

-   The widespread impact of COVID-19 events and vaccine innovation have been investigated in previous studies, but with less focus of the impact of these events on vaccine-producing companies themselves.

-   We focus on the three largest COVID-19 vaccine-producing companies by market capitalisation: Oxford-AstraZeneca, Janssen-Cilag International NV (sub-company of Johnson & Johnson), and Pfizer-BioNTech.

-   Pfizer-BioNTech was the first COVID-19 vaccine to be approved by the United States Food and Drugs Association.
:::

# Methods

## Data Collection

-   **Tools:** `yfR` (Yahoo Finance) R package.

-   **Time period:** January 30, 2015 to November 8, 2025

-   **Stocks:** Pfizer-BioNTech `(PFE)`, Oxford-AstraZeneca `(AZN)`, Johnson & Johnson `(JNJ)`

-   **Foreign exchange rates:** GBP to USD exchange rate `(GBPUSD)`

-   **Market & risk-free rates:** S&P 500 `(GSPC)`, 10-year US treasury yields `(TNX)`

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false
library(dplyr)
library(ggplot2)
library(tidyr)
library(janitor)
library(patchwork)

# all results
company_results <- readRDS("../output/company-specific-results.rds")

# 3 largest in market cap
file_paths <- c("../data/JNJ_yfR_2015-01-30_2025-09-08.rds",
                "../data/rds/AZN_USD_2015-01-30_2025-09-08.rds", 
                "../data/PFE_yfR_2015-01-30_2025-09-08.rds")
azn <- readRDS(file_paths[2])
jnj <- readRDS(file_paths[1])
pfe <- readRDS(file_paths[3])

# Function
cumret <- function(file_path) {
  # Read data
  df <- readRDS(file_path)

  price_close <- df$price_close
  ret <- price_close / dplyr::lag(price_close) -1
  cumret <- cumprod(1 + coalesce(ret, 0)) - 1

  # Preprocess
  df <- df %>%
    group_by(ref_date) %>%
    select(contains("ticker"), ref_date, price_close, volume, price_adjusted) %>%
    mutate(
      price_close = as.numeric(price_close),
      price_adjusted = as.numeric(price_adjusted),
      volume = as.numeric(volume)
    ) %>%
    arrange(ref_date) %>%
    mutate(
      ret = !!ret,
      cumret = !!cumret)
  
  min_cumret <- min(df$cumret)
  ticker_col <- names(df)[grepl("ticker", names(df))]
  ticker_name <- df[[ticker_col[1]]][1]
  
  # Return useful objects
  return(list(
    data = df,
    ret = ret,
    cumret = cumret
  ))
}

# combine data sets for stacked line chart
jnj_mod <- as.data.frame(cumret(file_paths[1])[[1]])
azn_mod <- as.data.frame(cumret(file_paths[2])[[1]])
pfe_mod <- as.data.frame(cumret(file_paths[3])[[1]])

combined_cumret <- bind_rows(jnj_mod, azn_mod, pfe_mod)
```

# Exploratory Data Analysis (EDA)

```{r}
#| eval: false
#| message: false
#| warning: false
#| include: false
#| paged-print: true

knitr::kable(head(azn), format = "markdown", caption = "Oxford-AstraZeneca (converted to USD)", digits = 2) %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = T,
                font_size = 13,
                position = "left")

knitr::kable(head(jnj), caption = "Johnson & Johnson", digits = 2)  %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = T,
                font_size = 13,
                position = "left")

knitr::kable(head(jnj), caption = "Pfizer-BioNTech", digits = 2)  %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = T,
                font_size = 13,
                position = "left")
```

## Data Description

-   We used time-series analysis on cumulative returns at time $i$ on each stock which was calculated from the **`price_adjusted`** column values.

    $$
    \text{cumulative return}_i = \frac{\text{adjusted price}_i - \text{adjusted price}_0}{\text{adjusted price}_i}
    $$

## 

-   **Events** were divided into two categories:

    -   **Market-wide**: events which were announced by international or governmental health organisations.

    -   **Firm-specific**: events which were announced or initiated by firms themselves.

## Market-wide events

::::: columns
::: {.column style="width "}
```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 7.5
#| paged-print: false

# plot stacked line chart - general events
gen_events <- c(as.Date("2020-01-30"),
                as.Date("2020-12-31"),
                as.Date("2023-05-05"))

zoomed_combined <- combined_cumret %>%
  mutate(
    ticker = ifelse(is.na(ticker), ticker.x, ticker)
  ) %>%
  filter(
    ref_date >= min(gen_events) - 300, 
    ref_date <= max(gen_events) + 300
  )

minc <- min(combined_cumret$cumret)

p_ge <- ggplot(zoomed_combined, aes(x = ref_date, y = cumret, group = ticker, color = ticker)) +
  geom_line() +
  geom_vline(
    xintercept = gen_events,
    colour = "gray30",
    linetype = "dashed"
    ) +
  annotate(
    "label",
    x = gen_events[1],
    y = 0.55,
    label = "PHEIC\nAnnounced",
    angle = 0,
    vjust = -0.5,
    size = 3,
    colour = "gray30"
  ) +
  annotate(
    "label",
    x = gen_events[2],
    y = 0.7,
    label = "First FDA\nApproval",
    angle = 0,
    vjust = -0.5,
    size = 3,
    colour = "gray30"
  ) +
  annotate(
    "label",
    x = gen_events[3],
    y = 0.8,
    label = "PHEIC\nRemoved",
    angle = 0,
    vjust = -0.5,
    size = 3,
    colour = "gray30"
  ) +
  labs(
    title = "Cumulative Returns of JNJ, AZN, and PFE",
    colour = "Ticker",
    x = "Date",
    y = "Cumulative returns since 2015-01-30"
  ) +
  theme_linedraw() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
p_ge

```
:::

::: {.column style="width "}
### Observations

-   Large dip in the months following official WHO pandemic announcement (PHEIC).

-   Effect of the first-ever FDA-approved COVID-19 vaccine is unclear.

-   Dip in all three stocks with removal of PHEIC status.
:::
:::::

## Firm-specific events

::::: columns
::: column
```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 6
#| paged-print: false
combined_data <- bind_rows(
  jnj_mod |> mutate(company = "JNJ"),
  azn_mod |> mutate(company = "AZN"),
  pfe_mod |> mutate(company = "PFE")
)

event_data <- tribble(
  ~company, ~date, ~label,
  "JNJ", as.Date("2020-07-10"), "Phase 1/2\nTrial Start",
  "JNJ", as.Date("2021-02-27"), "FDA EUA",
  "JNJ", as.Date("2021-04-13"), "Deployment\nPaused",
  "AZN", as.Date("2020-04-23"), "World Trials\nCommence",
  "AZN", as.Date("2020-09-06"), "US Trials Halt",
  "AZN", as.Date("2020-10-23"), "US Trials Resume",
  "AZN", as.Date("2021-03-07"), "First Clots",
  "AZN", as.Date("2021-03-16"), "More Clots",
  "PFE", as.Date("2020-11-01"), "UK EUA",
  "PFE", as.Date("2021-08-23"), "FDA Approval",
  "PFE", as.Date("2021-10-29"), "FDA EUA\nAges 5–11"
)

min_event <- min(event_data$date)
max_event <- max(event_data$date)

# zoom window: 20 days before earliest event to 60 days after latest (you can adjust)
zoomed_data <- combined_data %>%
  filter(
    ref_date >= min_event - 20,
    ref_date <= max_event + 60
  )

# --- Faceted plot ---
ggplot(zoomed_data, aes(x = ref_date, y = cumret)) +
  geom_line() +
  geom_vline(
    data = event_data,
    aes(xintercept = date),
    colour = "gray30",
    linetype = "dashed"
  ) +
  ggrepel::geom_label_repel(
    data = event_data,
    aes(x = date, y = Inf, label = label),
    colour = "gray30",
    fill = "white",
    label.size = 0.2,
    direction = "y",
    size = 2.5,
    force = 2
  ) +
  facet_wrap(~ company, scales = "free_y", ncol = 1) +
  labs(
    title = "Cumulative Returns Around Major COVID-19 Vaccine Events",
    x = "Date",
    y = "Cumulative Returns (since 2015-01-30)"
  ) +
  theme_linedraw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.background = element_rect(fill = NA)
  )

```
:::

::: column
### Observations

-   Possibly more activity during clinical trials and emergency use events 

    -   JNJ: Phase ½ trials

    -   AZN: US trials halt and result

    -   PFE: UK emergency use
:::
:::::

# Conducting Event Studies

## Selecting Statistical Models for Expected Return

-   **Market Model, Market Adjusted Model,** and **Comparison Period Mean Adjusted Model** are chosen primarily for firm-specific events.

    -   Market Model: $R_{i,t} = \alpha_i + \beta_i \cdot R_{m,t} + \epsilon_{i,t}$

    -   Market Adjusted Model: $R_{i,t} = R_{m,t} + \epsilon_{i,t}$

    -   Comparison Period Mean Adjusted Model: $R_{i,t} = \mu_i + \epsilon_{i,t}$

## 

-   **Fama-French 3 Factor Model** for market-wide events.

    $$
    R_i - R_f = \alpha + \beta_1 \cdot (R_m - R_f) + \beta_2 \cdot \text{SMB} +\beta_3 \cdot \text{HML} + \epsilon
    $$

    -   $R_i - R_f$: Actual return of stock

    -   $\alpha$: (intercept) Expected excess return of stock

    -   $R_m - R_f$: Market excess return

    -   **SMB (small minus big)**: Size premium

    -   **HML (high minus low)**: Value premium

    -   $\beta_{1,2,3}$: Est. coefficients for market, size, and value factors

    -   $\epsilon$: Abnormal returns

# Results

## Market-Wide Events

```{r}
fama <- readRDS("../output/fama_results.rds")

knitr::kable(fama, digits = 4) %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "bordered"),
    full_width = T,
    font_size = 24 
  ) %>%
  kableExtra::row_spec(
    row = 3,
    background = "darkseagreen2"
  ) %>%
  kableExtra::column_spec(
    column = 4,
    bold = ifelse(fama$p_event < 0.05, TRUE, FALSE)
  )
```

## Firm-Specific Events

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

significant_results <- company_results %>%
  select(ends_with("_p")) %>%
  mutate(across(everything(), ~ ifelse(. < 0.05, ., NA_real_))) %>%
  filter(if_any(everything(), ~ !is.na(.))) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), "", .)))

p_results <- company_results %>%
  select(ends_with("_p"))

knitr::kable(p_results, digits = 4) %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "bordered"),
    full_width = T,
    font_size = 24
  ) %>%
  kableExtra::row_spec(
    row = 1,
    background = "darkseagreen2"
  ) %>%
  kableExtra::column_spec(
    column = 2,
    bold = ifelse(p_results$MAM_p < 0.1, TRUE, FALSE)
  ) %>%
  kableExtra::column_spec(
    column = 3,
    bold = ifelse(p_results$CMA_p < 0.1, TRUE, FALSE)
  ) %>%
  kableExtra::column_spec(
    column = 4,
    bold = ifelse(p_results$MM_p < 0.1, TRUE, FALSE)
  ) %>%
  kableExtra::row_spec(
    row = 7,
    background = "darkseagreen2"
  ) %>%
  kableExtra::row_spec(
    row = 10,
    background = "darkseagreen2"
  ) %>%
  kableExtra::row_spec(
    row = 2,
    background = "lightgoldenrod1"
  ) %>%
  kableExtra::row_spec(
    row = 3,
    background = "lightgoldenrod1"
  ) %>%
  kableExtra::row_spec(
    row = 9,
    background = "lightgoldenrod1"
  )
```

## Conclusions

### Investor Behaviour

-   **Innovation** seemed to be a common factor in decision-making of investment into individual pharmaceutical companies.
-   Companies' market capitalisation and book-to-value ratio determine sensitivity to events.

### Methodology

-   Abnormality is model-dependent and *p*-values are interpreted differently under different models.

## Further Action

-   Explore how other firm-specific events impact external firms.

    -   ie. Pfizer-BioNTech Emergency Use Authorisation for 5-11 year olds by the US FDA seemed to cause a dip in all three stocks.

# Thank You! Questions?

## References

Perlin M (2023). *yfR: Downloads and Organizes Financial Data from Yahoo Finance*. R package version 1.1.0, https://github.com/ropensci/yfR.

Wickham H (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York. ISBN 978-3-319-24277-4, https://ggplot2.tidyverse.org.

World Health Organization (2023). *Status of COVID-19 Vaccines within WHO EUL/PQ evaluation process*. Guidance Document, https://extranet.who.int/pqweb/key-resources/documents/status-covid-19-vaccines-within-who-eulpq-evaluation-process.

## 

Nguyen M (2025). *A Guide on Data Analysis*. Bookdown, https://bookdown.org/mike/data_analysis/

Dal-Ré R (2023). *The winding 12-month journey of the AstraZeneca COVID-19 vaccine since its first administration to humans*. Therapie, 78(3), 293–302. https://doi.org/10.1016/j.therap.2022.07.003

Brice Y, Morgan L, Kirmani M, Kirmani M, & Udeh M C (2023). *COVID-19 Vaccine Evolution and Beyond.* Neuroscience insights, 18, 26331055231180543. https://doi.org/10.1177/26331055231180543

## 

Müller S (2023). *Choosing the Right Approach: A Comprehensive Guide to Estimating Expected Returns.* https://eventstudy.de/models/expected_return.html#economic-models

Wolf M, Schimmer M, Levchenko A, & Müller S (2014). *EventStudyTools (Research Apps)*, St.Gallen, http://www.eventstudytools.com.

Fama E F, French K R (2025). *Current Research Returns*. Data Library, https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html

## 

Martinez-Blasco M, Serrano V, Prior F & Cuadros J (2023). *Analysis of an event study using the Fama–French five-factor model: teaching approaches including spreadsheets and the R programming language*. Financ Innov 9, 76. https://doi.org/10.1186/s40854-023-00477-3
